Neural and Symbolic Spatiotemporal Optimization of Cosmological Particle-Mesh Simulation

### 2023-06-07 docs/papers/sto/so.tex
FIXUP draft: edit TODOs and text

\section*{TODO}
\begin{itemize}
\item (paper) Moving some of the following ongoing and completed tasks
  to the main text and add preliminary figures.
\item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
  different mass resolutions; probably replace linear snapshot spacing
  according to the nonlinearity (estimate (with PM till convergence) the
  std acc over the std vel) of each realization.
\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.\newline
    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
    consistent at the field level. To further check power spectrum. For odd mesh
    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
    slightly different. To figure out this weird issue.}
\item (model; YL) Probably necessary to first incorporate List \& Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\item (train, valid) Show $T - 1$ in symlog with 0.01 linthresh, and
  $1-r^2$ in log at 3 different $a$'s; Show $f$ and $g$ (along axis and
  face diagonal) at 3 different $a$'s; Track mean epoch loss
\item (train) Add random offset in loss grid (see \autoref{tab:param})
  to remove the position dependence in loss.
\item (train) Randomize snapshots of each simulation instead of
  sequential; alternatively accumulate their gradients
\item (train) 3 sets of grid: ptcl, force mesh, and loss grid. Explore
  if we can use a finer loss grid, from 1x to say 1.5x and even 2x.
\item (train, valid) Non-integer multiple mesh shapes: for uniform
  particle grid, test (with double precision on CPUs) whether they start
  moving by artifact forces on cell scales, in 3 cases: no
  deconvolution, with CIC deconvolution, with SO.
\item (train) For generalizability and robustness, consider testing
  L1/L2 and dropout. Also they're worth adding if they give lower losses
  at convergence. Note that 512 realization are not too many, and we
  have features that can be very correlated.
\item (train) find the best loss function for production training
\item (train) optuna hyperparam search on a subset of training data
  before production training
  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna example}).
\item (train) Reduce learning rate on (mean epoch) loss plateau in
  production
\item (valid) Interpolation test: compare SO DM and halo fields and
  summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
  on the last 9 test cases and noncubic box.
\item (valid) Extrapolation test: demonstrate generalization capability,
  compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
\end{itemize}

@@

@@

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \cL_\vs + \cL_\vv + \cL_\delta
+ \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
+ \cdots
\end{equation}
%
\YL{(a simple sum since we have already done careful normalization in
$\cL_f$'s) and also try}
%
\begin{equation}
\cJ = \ln\cL_\vs + \ln\cL_\vv + \ln\cL_\delta
+ \ln\cL_{(1+\delta)\vv} + \ln\cL_{(1+\delta)\vv\vv}
+ \cdots
\end{equation}
%
Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
Lagrangian space, while the rest are in the Eulerian space:
%
\begin{align}
1 + \delta(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
%
\bigl[(1+\delta)\vv\bigr]_i(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i, \nonumber\\
%
\bigl[(1+\delta)\vv\vv\bigr]_{ij}(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i v_j,
\end{align}
%
which we compute by scattering particles and their velocities to the
\YL{choice of mesh for loss}.

An alternative loss function can be the Eulernian density weighted
particle displacement MSE.

@@

In Fourier space, the gravitational force field reads
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
Force sharpening
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$, $g$ are neural networks, each approximating a nonlinear
function of the wavevector components and all the features considered
above in dimensionless combinations, e.g.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}
%
$f$ and $g$ are optimized during training to correct and sharpen the
gravitational forces.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ can be the most general way to modify the
force kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

@@

Given 512 \GADGET\ simulations and 121 output snapshots each, in total we have
61952 snapshots for training.
Following \autoref{tab:param}, given a snapshot from \GADGET\ simulations, we
configure \pmwd\ with the same initial coniditions and cosmological parameters,
and randomly sample the mesh shape (or the mesh cell size) and the number of
time steps.

We perform parallel training using a number of nodes that each hosts multiple
GPU devices.
The training on a given snapshot is performed on a single GPU device.
The gradients are collected and averaged to update the SO parameters.
To make sure that each individual training in a parallel run finishes in similar
time, they share the same number of time steps.

>>>>

\section*{TODO}
\begin{itemize}
\item (paper) Moving some of the following ongoing and completed tasks
  to the main text and add preliminary figures.
\item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
  different mass resolutions; probably replace linear snapshot spacing
  according to the nonlinearity (estimate (with PM till convergence) the
  std acc over the std vel) of each realization.
\item (model; YL) Change nbody for loop by lax scan and/or while loops.
  while\_loop can help to save compilation when trained on different
  snapshots. Performance difference between scan and while?
\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.\newline
    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
    consistent at the field level. To further check power spectrum. For odd mesh
    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
    slightly different. To figure out this weird issue.}
\item (model; YL) Probably necessary to first incorporate List \& Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\item (train, test) For $\vs(\vq)$, $\vv(\vq)$, $\delta(\vx)$,
  $(1+\delta)\vv(\vx)$, $(1+\delta)\vv\vv(\vx)$, show $T - 1$ in symlog
  with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
  and $g$ (along axis and face diagonal) at 3 different $a$'s; Track
  mean epoch loss
\item (train) Randomize snapshots of each simulation instead of
  sequential; alternatively accumulate their gradients; Normalize loss
  by number of time steps
\item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
  random offset in loss grid (see \autoref{tab:param}) to remove the
  position dependence in loss; likewise for force mesh? how about only
  training (for lower loss), only ``inference'' (for higher accuracy),
  or both? the random force mesh offset can also be replaced by a
  structured scheme, e.g., alternating between 2 interlocking grids
\item (train) Explore if we can use a finer loss grid, from 1x to say
  1.5x and even 2x.
\item (train, test) Non-integer multiple mesh shapes: for uniform
  particle grid, test (with double precision on CPUs) whether they start
  moving by artifact forces on cell scales, in 3 cases: no
  deconvolution, with CIC deconvolution, with SO.
\item (train) find the best loss function for production training
\item (train) For generalizability and robustness, consider testing
  weight decay and dropout. Also they're worth adding if they give lower
  losses at convergence. Note that 512 realization are not too many, and
  we have features that can be very correlated.
\item (train) optuna hyperparam search on a subset of training data
  before production training
  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna example}).
\item (train) Reduce learning rate on (mean epoch) loss plateau in
  production
\item (test) Interpolation test: compare SO DM and halo fields and
  summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
  on the last 9 test cases and noncubic box.
\item (test) Extrapolation test: demonstrate generalization capability,
  compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
\end{itemize}

@@

\YL{
For a field $f$ in either Lagrangian or Eulerian space, we have thought
of 3 forms of MSE-inspired loss:
%
\begin{equation}
\cL_f = \ln \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
                   {\sum_{\vq'} f(\vq')^2},
\end{equation}
%
(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
their Fourier wavevector $\vk$ in both cases)
%
\begin{equation}
\cL_f = \frac1{N_K} \sum_i \ln
\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} | f(\vk')|^2},
\end{equation}
%
and
%
\begin{equation}
\cL_f = \ln \biggl[ \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
\end{equation}
%
Then the question is which loss we should use for which $f$, to get
stable and robust results, e.g., no bump for large or small boxes.
To further limit our scope in this paper, we should probably only
consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
velocity $\vv(\vq)$.
}

@@

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \frac1{N_a} (\cL_\vs + [\cL_\vv] + \cL_\delta),
\end{equation}
%
where $N_a$ is the time steps it take to generate the snapshot for each
iteration.
\YL{(a simple sum since we have already done careful normalization in
$\cL_f$'s) and also try}
%
\begin{equation}
\cJ = \frac1{N_a} (\ln\cL_\vs + [\ln\cL_\vv] + \ln\cL_\delta).
\end{equation}
%

\YL{TODO move this above:}
Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
Lagrangian space, while the density field is in the Eulerian space:
%
\begin{equation}
1 + \delta(\vx) = \int\!\d^3\vq\, \deltaD[ \vx - \vq - \vs(\vq)],
\end{equation}
%
which we compute by scattering particles to the choice of grid for loss.

@@

In Fourier space, the gravitational force field reads
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
Force sharpening
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$, $g$ are neural networks, each approximating a nonlinear
function of the wavevector components and all the features considered
above in dimensionless combinations, e.g.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}
%
$f$ and $g$ are optimized during training to correct and sharpen the
gravitational forces.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

@@

Given 512 \GADGET\ simulations and 121 output snapshots each, in total
we have 61952 snapshots for training.
Following \autoref{tab:param}, in each iteration, we randomly sample the
snapshots, the number of time steps, and the mesh shape (or its cell
size), and configure \pmwd\ with the same initial coniditions,
cosmological parameters and other common configurations.
To balance the work of GPUs during training, in each iteration we use
the same snapshot $a$, number of time steps, and mesh shape on all GPU.

We perform parallel training using a number of nodes that each hosts
multiple GPU devices.
The training on each snapshot is performed on a single GPU device.
The gradients are collected and averaged to update the SO parameters.

Hyperparameters:

### 2023-06-07 docs/papers/sto/sto.tex
FIXUP draft: loss norm caveat
FIXUP draft: adding loss norm inside log
FIXUP draft: edit TODOs and text

\author[Zhang, Li, Jamieson, et al.]{
%
Yucheng Zhang,$^{1, 2}$
%
Yin Li,$^{1, 3, 4}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
%
Drew Jamieson,$^{5}$
%
Libin Lu,$^{3}$
%
and XYZ
%
\\$^1$Department of Mathematics and Theory, Peng Cheng Laboratory,
Shenzhen, Guangdong 518066, China
%
\\$^2$Center for Cosmology and Particle Physics, Department of Physics,
New York University, New York, NY 10003, USA
%
\\$^3$Center for Computational Mathematics, Flatiron Institute, New
York, NY 10010, USA
%
\\$^4$Center for Computational Astrophysics, Flatiron Institute, New
York, NY 10010, USA
%
\\$^5$Max Planck Institute for Astrophysics, 85748 Garching bei
M\"unchen, Germany
}

@@

\begin{abstract}
We sharpen PM force and optimize integration time steps.
\end{abstract}

@@

\section*{TODO}
\begin{itemize}
\item Separate sto to SO and TO.
\item Moving some of the following ongoing and completed tasks to the
    main text and add preliminary figures.
\item Implement the rest of the nonlinear scales, including the rms
    displacement with np.trapz of $P$ over $\ln k$, the $\ln a$ time
    derivatives, etc.\newline
    \textbf{Update}: code has been added.
\item Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.\newline
    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
    consistent at the field level. To further check power spectrum. For odd mesh
    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
    slightly different. To figure out this weird issue.}
\item Be careful with the non-deterministic ICs generated on GPUs.\newline
      \textbf{Update}: we have shown that ICs generated on GPUs are
      deterministic (but different from those on CPUs), generating ICs for both
      training data and training on GPUs might be the simplest solution.
\item Incorporate 3lpt branch; testing; find a\_start for different mass
  resolutions
\item Probably necessary to first incorporate List \& Hahn's recent
  improvement on FastPM integrator arXiv:2301:09655.
\item Compare sto DM and halo fields and summary stats with those of raw
  pmwd, deconvolved pmwd, and \GADGET, on the last 9 test cases and
  noncubic box.
\item Demonstrate capability of generalization, on some cases with
  extrapolating cosmological parameters, etc.
\item Non-integer multiple mesh shapes: for uniform particle grid, test
  (with double precision on CPUs) whether they start moving by artifact
  forces on cell scales, in 3 cases: no deconvolution, with CIC
  deconvolution, with SO.
\item In TO, time step spacing can be important for light cone
  generation. We can train on light cone target data to learn TO for
  light cones. The result may be related to $c \d t/H = c \d a / a^2 H$.
\end{itemize}

@@

For the displacement field, we adopt the mean squared error (MSE) loss
that proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2,
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es) below.
% Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too

However, in the Eulerian space, an MSE loss is dominated by the small
scale modes.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
Our loss on the overdensity field $\delta$ sums up its MSE relative to
the target field power in logarithmic bins:
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i \,\ln\!\!
\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2,
\end{equation}
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |\delta(\vk')|^2},
\end{equation}
%
where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
factor that suppresses the high $k$ contribution in Eulerian fields such
as the density field (with $w=1$ for the Lagrangian displacement field).
This is because a PM solver gradually loses small scale information step
by step, and without $w$ setting the priority, the loss function could
encourage the network to fit the small scales at the cost of the
intermediate scales.
As a result, we can set the loss mesh size the same as that of the PM
force solver.
We can try for $w$:
%
\begin{equation}
\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{to some power},
\prod_{j=1}^3 \cos (k_j l / 2) \textrm{to some power},
\end{equation}
%
which vanish when any $k_j$ reaches the Nyquist.
\YL{Maybe add a figure of $w$ along axis, face diagonal and body
digonal, if this works.}

To understand the form of the loss function, let's introduce for each
mode the square root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
\end{equation}
%
and the cross correlation coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
\end{equation}
%
Now, under the mild assumptions that $T$ and $r$ are smooth and
isotropic, we can show
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, \ln
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
\end{equation}
%
up to a constant;
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
\end{equation}
%
that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
when $\hat\delta$ is identical to $\delta$.
Above we choose the binned version instead for the robustness against
the noisy modewise $|\delta(\vk)|^2$ in the denominator.
The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
loss, while our modification is important to balance among all scales so
that the neural network does not focus too much on either high or low
$k$'s.

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \cL_\vs + \cL_\delta
+ \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
+ \cdots
\end{equation}
%
\YL{(a simple sum since we have already done careful normalization in
$\cL_f$'s) and also try}
%
\begin{equation}
\cJ = \ln \cL_\vs + \ln \cL_\delta
+ \ln \cL_{(1+\delta)\vv} + \ln \cL_{(1+\delta)\vv\vv}
+ \cdots
\end{equation}
%
Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
Lagrangian space, while the rest are in the Eulerian space:
%
\begin{align}
1 + \delta(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
%
\bigl[(1+\delta)\vv\bigr]_i(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i, \nonumber\\
%
\bigl[(1+\delta)\vv\vv\bigr]_{ij}(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i v_j,
\end{align}
%
which we compute by scattering particles and their velocities to the
\YL{choice of mesh for loss}.

An alternative loss function can be the Eulernian density weighted
particle displacement MSE.

@@

@@

In Fourier space, the gravitational force field reads
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
Force sharpening
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
\end{equation}
%
where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
function of the dimensionless features, e.g.,
%
\begin{equation}
g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}

Parameters in these neural networks are optimized during training.

\YL{While $g$ is spherically symmetric by assumption, $f$ and $h$ are
expected to be even functions, i.e., real valued in Fourier space.
It is interesting to symbolically regress both $h(k_i)$ and $k_i h(k_i)$
to see which one gives simpler expression, given the possible different
rankings by complexity.
For example, $k_i h(k_i)$ may have a simpler form as a nonlinear odd
function.}

@@

Given 512 \GADGET\ simulations and 121 output snapshots each, in total we have
61952 snapshots for training.
Following \autoref{tab:param}, given a snapshot from \GADGET\ simulations, we
configure \pmwd\ with the same initial coniditions and cosmological parameters,
and randomly sample the mesh shape (or the mesh cell size) and the number of
time steps.

We perform parallel training using a number of nodes that each hosts multiple
GPU devices.
The training on a given snapshot is performed on a single GPU device.
The gradients are collected and averaged to update the STO parameters.
To make sure that each individual training in a parallel run finishes in similar
time, they share the same number of time steps.

>>>>

\author[Zhang \& Li et al.]{
%
Yucheng Zhang,$^{1, 2}$
%
Yin Li,$^{1, 3, 4}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
%
Drew Jamieson,$^{5}$
%
Libin Lu,$^{3}$
%
and XYZ
%
\\$^1$Department of Mathematics and Theory, Peng Cheng Laboratory,
Shenzhen, Guangdong 518066, China
%
\\$^2$Center for Cosmology and Particle Physics, Department of Physics,
New York University, New York, NY 10003, USA
%
\\$^3$Center for Computational Mathematics, Flatiron Institute, New
York, NY 10010, USA
%
\\$^4$Center for Computational Astrophysics, Flatiron Institute, New
York, NY 10010, USA
%
\\$^5$Max Planck Institute for Astrophysics, 85748 Garching bei
M\"unchen, Germany
}

@@

\begin{abstract}
We optimize integration time steps with first neural networks and then
symbolic regression, taking dimensional analysis in consideration.
\end{abstract}

@@

\section*{TODO}
\begin{itemize}
\item In TO, time step spacing can be important for light cone
  generation. We can train on light cone target data to learn TO for
  light cones. The result may be related to $c \d t/H = c \d a / a^2 H$.
\end{itemize}

@@

For the displacement field, we adopt the mean squared error (MSE) loss
that proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
                   {\sum_{\vq'} s(\vq')^2},
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es) below.
% Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too

\YL{Even though the log should be enough for adaptive optimizer and
fixed sim config, this may create problem when training on all config at
the same time. So normalizations inside the log is added above and
below.}

However, in the Eulerian space, an MSE loss is dominated by the small
scale modes.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
Our loss on the overdensity field $\delta$ sums up its MSE relative to
the target field power in logarithmic bins:
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i \ln
\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |\delta(\vk')|^2},
\end{equation}
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |\delta(\vk')|^2},
\end{equation}
%
where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
factor that suppresses the high $k$ contribution in Eulerian fields such
as the density field (with $w=1$ for the Lagrangian displacement field).
This is because a PM solver gradually loses small scale information step
by step, and without $w$ setting the priority, the loss function could
encourage the network to fit the small scales at the cost of the
intermediate scales.

To understand the form of the loss function, let's introduce for each
mode the square root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
\end{equation}
%
and the cross correlation coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
\end{equation}
%
Now, under the mild assumptions that $T$ and $r$ are smooth and
isotropic, we can show
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, \ln
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
\end{equation}
%
replacing
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
\end{equation}
%
that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
when $\hat\delta$ is identical to $\delta$.
Above we choose the binned version instead for the robustness against
the noisy modewise $|\delta(\vk)|^2$ in the denominator.
The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
loss, while our modification is important to balance among all scales so
that the neural network does not focus too much on either high or low
$k$'s.

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \frac1{N_a} (\cL_\vs + \cL_\vv + \cL_\delta
+ \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
+ \cdots),
\end{equation}
%
where $N_a$ is the time steps it take to generate the snapshot for each
iteration.
\YL{(a simple sum since we have already done careful normalization in
$\cL_f$'s) and also try}
%
\begin{equation}
\cJ = \frac1{N_a} (\ln\cL_\vs + \ln\cL_\vv + \ln\cL_\delta
+ \ln\cL_{(1+\delta)\vv} + \ln\cL_{(1+\delta)\vv\vv}
+ \cdots).
\end{equation}
%
Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
Lagrangian space, while the rest are in the Eulerian space:
%
\begin{align}
1 + \delta(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
%
\bigl[(1+\delta)\vv\bigr]_i(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i, \nonumber\\
%
\bigl[(1+\delta)\vv\vv\bigr]_{ij}(\vx) &= \int\!\d^3\vq\,
\deltaD[ \vx - \vq - \vs(\vq)] v_i v_j,
\end{align}
%
which we compute by scattering particles and their velocities to the
choice of grid for loss.

@@

\begin{table}
\centering
\caption{Possible features that can affect force sharpening. TODO}
\label{tab:feat}
\begin{tabular}{rl}
\toprule
type & feature \\
\midrule
& $R_\mathrm{d}$: RMS linear displacement \\
\cmidrule(lr){2-2}
& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
\cmidrule(lr){2-2}
nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $\d R / \d\lna$: time derivatives of the above \\
\midrule
& particle spacing \\
\cmidrule(lr){2-2}
other scales & $l$: PM force mesh cell size \\
\cmidrule(lr){2-2}
& softening length \\
\midrule
& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
\cmidrule(lr){2-2}
& $\d\ln G_m / \d\lna$ \\
\cmidrule(lr){2-2}
dimensionless factors & $\Omegam(a)$ \\
\cmidrule(lr){2-2}
& $\d\ln\!H / \d\lna$ \\
\cmidrule(lr){2-2}
& $\Delta\lna$: time step size in $\lna$ \\
\bottomrule
\end{tabular}
\end{table}

@@

In Fourier space, the gravitational force field reads
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
Force sharpening
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$, $g$ are neural networks, each approximating a nonlinear
function of the wavevector components and all the features considered
above in dimensionless combinations, e.g.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}
%
$f$ and $g$ are optimized during training to correct and sharpen the
gravitational forces.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
etc to see which one gives simpler expression, given the possible
different rankings by complexity.
For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
function.}

@@

Given 512 \GADGET\ simulations and 121 output snapshots each, in total
we have 61952 snapshots for training.
Following \autoref{tab:param}, in each iteration, we randomly sample the
snapshots, the number of time steps, and the mesh shape (or its cell
size), and configure \pmwd\ with the same initial coniditions,
cosmological parameters and other common configurations.
To balance the work of GPUs during training, in each iteration we use
the same snapshot $a$, number of time steps, and mesh shape on all GPU.

We perform parallel training using a number of nodes that each hosts
multiple GPU devices.
The training on each snapshot is performed on a single GPU device.
The gradients are collected and averaged to update the SO parameters.

Hyperparameters:
